\vspace{-.2in} Squint: Lossy Hierarchical Compression on Symbol Streams \vspace{-.1in}

### 2012-12-11 tex/paper.tex
Fixing up intro and formatting title page

\newcommand{\Sequitur}{\textsc{Sequitur}\xspace}

@@

\title{Squint: Lossy Hierarchical Compression on Symbol Streams}
\author{Edward Gan \& Max Wang}
\date{December 11, 2012}
\maketitle

@@

Grammar-based compression schemes like \Sequitur exploit repeated structures in
streams of symbolic data to achieve good compression ratios.  We develop a set
of algorithms for improving upon Grammar-based compression schemes by allowing
some loss in the compression. Although precise subsequences might be altered,
in return we can simplify the grammars produced to obtain better ratios while
still preserving important syntatic patterns at both the micro and macro
levels.

@@

Many sources of data exhibit natural hierarchical structure. For instance,
english text often can be grouped well into sections, paragraphs, sentences,
...  all the way down to words and phonemes. Music too, consists of repeated
motifs, phrases, themes, and so on. Finally, computer code is so hierarchical
that its syntax can usually be put in BNF. This allows grammar-based
compression schemes to build up Context Free Grammars (CFGs) and achieve very
good compression ratios.

Taking a step back, in the image compression world there are both lossless and
lossy compression schemes. In certain situations, such as for low-pixel count
logos, one could use a lossless image compression scheme. In other situations
however, such as for textures and some photographs, the exact pixel values are
almost irrelevant compared to the overall gradients, colors, and structures in
the image, and one could use a lossy compression scheme like jpeg.

We believe it is valuable to bring some of the attitudes prevalent in image/
video compression to the world of hierarchical text compression. Better
compression ratios for text are possible.

It may seem strange to allow a string such as "mind" to be lossily corrupted
into ``rind'', but if one is more interested in recurrent structure in the text
than in its precise meaning, these changes aren't important. For instance, if
one is looking at a text in a foreign language, one is more likely to pick up
the texture, the repeated sounds, some phrases that come up and the breaks
between larger scale sections.  A change in a single letter probably wouldn't
be noticed. Similarly, when skimming through huge files, perhaps execution
logs, one usually first starts looking for patterns or things that stand out.
Lorem ipsum is a perfect example of a piece of ``text'' whose value lies in its
texture and sentence/paragraph structure, and we will come back to it again.

In this paper we develop 2 algorithms and implement a system for lossily
compressing text while preserving its hierarchical structures. The goal was to
preserve the feel of a text at both micro and macro levels.  Since
grammar-based compression schemes were designed to capture these structures, we
built our system on top of the existing \Sequitur grammar inference algorithm.
Starting with the CFG generated by a lossless grammar inference algorithm, we
repeatedly simplify it, throwing out details along the way to emphasize
repetitions. Our main contribution lies in the development, implementation, and
evaluation of 2 algorithms for ``simplifying'' CFG grammars while preserving
their structure.

@@

The previous algorithm updates the grammar sequentially, since each variable
replacement can have an effect on large parts of the grammar, we will have to
go through comparing all pairs again after the change.  An alternative
algorithm would be to try to replace similar variables in parallel. This would
avoid many of the repeat pair-comparisons done in \emph{Similarity}. In other
words, the algorithm below puts similar variables into equivalence class
clusters, and then replaces each variable in a cluster them with the smallest
representative all at once each iteration.

@@

Suppose for the sake of contradiction that the Similarity algorithm created a
cycle at one step by redirecting an edge originally pointing at $A$ to $B$.
This would mean that there was a path going from $B$ to $A$. However, by
comparing the sizes of $A$ and $B$ we know that $A$ must have been bigger than
$B$. Thus there cannot have been a path from $B$ to $A$, and we have a
contradiction.

For the Cluster algorithm the analysis is a bit more involved, since to avoid
recomputation we only calculate the sizes of the variables once before doing
all of the replacements. First note that even as we modify the DAG, all of the
edges we create in the DAG will respect the ordering of sizes that we
calculated before we did any replacements. This is because each time we
redirect an edge from $X\rightarrow A$ to $X\rightarrow B$
$size(X)>size(A)>size(B)$ and transitivity will preserve the ordering.

Suppose then that we created a cycle during Cluster by redirecting an edge from
$X \rightarrow A$ to $X \rightarrow B$.  This means that there must have been a
path from $B$ to $X$. However, even in its intermediate state we know that the
DAG respects the $size$ ordering, and $size(X) > size(A) > size(B)$, so there
cannot be a path from $B$ to $X$. This is a contradiction and we are done.

@@

Given an input of length $n$, let $r$ be the number of rules generated by
\Sequitur from the input.  In the analysis of the Similarity algorithm, let $R$
be the total number of variable replacements we perform. For each replacement,
we had to have searched though up to all pairs of variables, and for each pair
of variables expanding them and calculating the levenshtein distance could take
time linear in $n$.  Performing the replacement only takes time linear in $r$.
Thus the runtime is $\boxed{O(R*r^2*n)}$.

@@

The inner loops of the Cluster are similar to those of the Similarity
algorithm.  We iterate the entire clustering and replacement procedure $C$
times. Going through each variable-cluster pair takes up to $O(r^2)$ time and
calculating the levenshtein distance takes $O(n)$ time. The time taken to
actually perform the replacements each iteration is absorbed into the $O(r^2)$
term. So, again we have a runtime of $\boxed{O(C*r^2*n)}$. Just as with the
similarity algorithm, a variety of factors make the runtime much better than
the $\boxed{O(n^4)}$ worst case scenario.  The big gain over the similarity
algorithm is that $C$ turns out to be nearly constant, since there are very few
huge chains of reductions that are only possible after other reductions have
been performed, an in fact we are able to perform a very large number of the
remaining reductions each time we perform an iteration of the clustering
algorithm. Thus a more reasonable upper bound for the runtime of cluster would
be $\boxed{O(n^3)}$

@@

When comparing grammars, we chose to define the \emph{size} of a grammar as the
sum of the lengths of the rhs's of the production rules.  \Sequitur yields a
relatively complicated inferred grammar with 13 rules and a size of 51. Both
the Similarity and Clustering Algorithms with $\epsilon = .4$ , yield the
following grammar after reduction. It has 5 rules and is almost half the size,
at 30.

@@

For our first suite of tests, we compare the effectiveness of the
\emph{Similarity} and \emph{Cluster} algorithms. We might expect Similarity to
achieve better quality at the same compression ratio since it updates the state
of the grammar between each variable replacement, on the other hand since it
must update the grammar it has the potential to run much more slowly than
Cluster.

@@

In table \ref{simclus_r_t} we report the results on running \Sequitur and
Cluster, without the grammar reductions, on the rep-n datasets. Time is the
wall clock running as measured on a 2.26 Ghz Intel Core 2 Duo Macbook Pro.
Ratio is the relative size of the grammar produced compared with \Sequitur,
where the size of the grammar is the sum of the lengths of the rhs's of the
rules. Levdistance is the levenshtein distance between the output of the
lossified grammar and the original input string, it is a kind of measure of the
fidelity of our algorithms but doesn't take into account how well patterns were
preserved. For the rep-n tests, proportion is the proportion of the final
string produced that was `*'. The closer to 1/30~0.033, the more accurate the
lossifier was in recreating the original texture of the input.

The columns of the data are prefixed by S or T depending on whether they are
measurements for Similarity or Cluster respectively.

@@

The actual compression ratio was respectable, with both Similarity and Cluster
coming in at nearly half the size of the \Sequitur CFG.  Setting higher values
for $\epsilon$ had diminishing returns for compression for larger and larger
file sizes, at the cost of drastically lowering the fidelity of the output.

@@

The differences between the two algorithms lie almost entirely in their
runtime. Both algorithms perform much worse on the less structued gen-n data,
we believe because simplifications are made more incrementally and the grammar
does not get dramatically simpler after a few iterations of our algorithm like
it does for the rep-n data.  Unfortunately the Similarity algorithm tends to
consistently incur an overhead of about $O(\sqrt{n})$ on top of the runtime for
the Cluster algorithm. Because of this we dismiss the Similarity algorithm as
unscalable and do not test it further.

@@

Though we were ultimately not able to come up with a good way to quantify how
much of the structure of a text was preserved, we have included the final
output strings we obtain after running the Clustering Algorithm with $\epsilon
= .4$ in the Appendices. We believe these tests serve as a good (subjective)
summary of the strengths and weaknesses of our algorithm in compressing data
while preserving its character.

The first piece of text is from the gen-n dataset again, the first 256000
characters of the book of Genesis. As listed in table \ref{simclus_g_t}, the
compressed grammar was size 5540 compared with \Sequitur's size of 8060.
Without knowledge of English spelling, the Cluster algorithm produces output
that is hard to read. However, key words are clearly recognizable, especially
important words such as `God', and the overall structure in the book of Genesis
is retained.

>>>>

\newcommand{\Sequitur}{\textsc{Sequitur}\xspace}
\newcommand{\Similarity}{\textsc{Similarity}\xspace}
\newcommand{\Cluster}{\textsc{Cluster}\xspace}

@@

\title{%
  \vspace{-.2in}%
  Squint: Lossy Hierarchical Compression on Symbol Streams%
  \vspace{-.1in}%
}
\author{Edward Gan \& Max Wang}
\date{%
  \vspace{-.1in}%
  December 11, 2012%
}

\maketitle

@@

Grammar-based compression schemes like \Sequitur exploit repeated structures in
streams of symbolic data to achieve good compression ratios.  We develop a set
of algorithms for improving upon grammar-based compression schemes by allowing
some lossiness in the compression.  Although precise subsequences might be
altered, in return we can simplify the grammars produced to obtain better
ratios while still preserving important syntatic patterns at both the micro and
macro levels.

@@

Many sources of data exhibit natural hierarchical structure.  For instance,
English text often can be sensibly grouped into sections, paragraphs,
sentences, and so on, all the way down to words and phonemes.  Likewise, music
consists of repeated motifs, phrases, themes, etc., which make up musical
structure.  Programming languages are so hierarchical that their syntax can
usually be represented in BNF form.  It is natural, then, to represent these
types of data using Context-Free Grammars (CFGs), which capture these
hierarchical structures; this form of \emph{grammar-based compression} can be
used to achieve very good compression ratios.

In the parallel world of image compression, however, two different classes of
compression schemes exist---\emph{lossless compression} and \emph{lossy
compression}.  In lossy compression schemes, rather than representing data
exactly, we allow a certain amount of corruption in order to achieve better
compression ratios.  Some situations may demand lossless encoding---say, in the
case of a corporation's trademarked graphical signature.  Many other
situations, however, may have less stringent requirements.  For texture
patterns and some photographs, the exact pixel values are far less important
than the overall gradients, colors, and structures in the image.  In such
cases, lossy compression schemes, such as JPEG for images \cite{jpeg}, can be
valuable and effective.

We believe it is worthwhile to introduce these notions of lossy compression
that are prevalent in image/video compression into the world of hierarchical
text compression.  Introducing lossiness to text compression opens up the
possibility for better compression ratios.

It may seem strange at first to allow a string such as "mind" to be lossily
corrupted into, say, ``rind''.  If, however, one is more interested in
recurrent structure in the text than in its precise semantic meaning, these
corruptions are unimportant.  For instance, when looking at text in a foreign
language, one is more likely to pick up the texture of the passage, the
repeated sounds, common phrases, and breaks between larger-scale sections of
text.  In this situation, a change of a single letter would probably go
unnoticed.  Similarly, when skimming through a huge file, say, an execution
log, an administrator usually begins by looking for patterns that stand out.
Lorem ipsum is a perfect example of a piece of ``text'' whose value lies in its
texture and sentence/paragraph structure, and we will return to it in later
sections.

In such situations, lossy hierarchical compression schemes may prove beneficial
if they can provide better compression ratios.  Moreover, introducing lossiness
to the CFGs which represent the input stream may accentuate more important
parts of the stream's structure.  Compression of a musical passage might, for
instance, capture important motives in favor of showing every rhythm and note.
Similarly, compression of text in a particular language may highlight
syntactical features.  We consider this a relevant potential use of lossy
hierarchical compression.

In this paper, we develop two algorithms and implement a broader system for
lossily compressing text while preserving its hierarchical structures.  The
goal was to preserve the feel of the text at both micro and macro levels.
Since grammar-based compression schemes were designed to capture these
structures, we built our system on top of the existing \Sequitur grammar
inference algorithm.  Starting with the CFG generated by a lossless grammar
inference algorithm, we repeatedly simplify it, throwing out details along the
way to emphasize repetitions.  Our main contribution lies in the development,
implementation, and evaluation of two algorithms for ``simplifying'' CFG
grammars while attempting to preserve their structure.

@@

The previous algorithm updates the grammar sequentially, since each variable
replacement can have an effect on large parts of the grammar, we will have to
go through comparing all pairs again after the change.  An alternative
algorithm would be to try to replace similar variables in parallel. This would
avoid many of the repeat pair-comparisons done in \Similarity. In other
words, the algorithm below puts similar variables into equivalence class
clusters, and then replaces each variable in a cluster them with the smallest
representative all at once each iteration.

@@

Suppose for the sake of contradiction that the \Similarity algorithm created a
cycle at one step by redirecting an edge originally pointing at $A$ to $B$.
This would mean that there was a path going from $B$ to $A$. However, by
comparing the sizes of $A$ and $B$ we know that $A$ must have been bigger than
$B$. Thus there cannot have been a path from $B$ to $A$, and we have a
contradiction.

For the \Cluster algorithm the analysis is a bit more involved, since to avoid
recomputation we only calculate the sizes of the variables once before doing
all of the replacements. First note that even as we modify the DAG, all of the
edges we create in the DAG will respect the ordering of sizes that we
calculated before we did any replacements. This is because each time we
redirect an edge from $X\rightarrow A$ to $X\rightarrow B$
$size(X)>size(A)>size(B)$ and transitivity will preserve the ordering.

Suppose then that we created a cycle during \Cluster by redirecting an edge from
$X \rightarrow A$ to $X \rightarrow B$.  This means that there must have been a
path from $B$ to $X$. However, even in its intermediate state we know that the
DAG respects the $size$ ordering, and $size(X) > size(A) > size(B)$, so there
cannot be a path from $B$ to $X$. This is a contradiction and we are done.

@@

Given an input of length $n$, let $r$ be the number of rules generated by
\Sequitur from the input.  In the analysis of the \Similarity algorithm, let $R$
be the total number of variable replacements we perform. For each replacement,
we had to have searched though up to all pairs of variables, and for each pair
of variables expanding them and calculating the levenshtein distance could take
time linear in $n$.  Performing the replacement only takes time linear in $r$.
Thus the runtime is $\boxed{O(R*r^2*n)}$.

@@

The inner loops of the \Cluster are similar to those of the \Similarity
algorithm.  We iterate the entire clustering and replacement procedure $C$
times. Going through each variable-cluster pair takes up to $O(r^2)$ time and
calculating the levenshtein distance takes $O(n)$ time. The time taken to
actually perform the replacements each iteration is absorbed into the $O(r^2)$
term. So, again we have a runtime of $\boxed{O(C*r^2*n)}$. Just as with the
similarity algorithm, a variety of factors make the runtime much better than
the $\boxed{O(n^4)}$ worst case scenario.  The big gain over the similarity
algorithm is that $C$ turns out to be nearly constant, since there are very few
huge chains of reductions that are only possible after other reductions have
been performed, an in fact we are able to perform a very large number of the
remaining reductions each time we perform an iteration of the clustering
algorithm. Thus a more reasonable upper bound for the runtime of cluster would
be $\boxed{O(n^3)}$

@@

When comparing grammars, we chose to define the \emph{size} of a grammar as the
sum of the lengths of the rhs's of the production rules.  \Sequitur yields a
relatively complicated inferred grammar with 13 rules and a size of 51. Both
the \Similarity and \Cluster algorithms with $\epsilon = .4$ , yield the
following grammar after reduction. It has 5 rules and is almost half the size,
at 30.

@@

For our first suite of tests, we compare the effectiveness of the
\Similarity and \Cluster algorithms. We might expect \Similarity to
achieve better quality at the same compression ratio since it updates the state
of the grammar between each variable replacement, on the other hand since it
must update the grammar it has the potential to run much more slowly than
\Cluster.

@@

In table \ref{simclus_r_t} we report the results on running \Sequitur and
\Cluster, without the grammar reductions, on the rep-n datasets. Time is the
wall clock running as measured on a 2.26 Ghz Intel Core 2 Duo Macbook Pro.
Ratio is the relative size of the grammar produced compared with \Sequitur,
where the size of the grammar is the sum of the lengths of the rhs's of the
rules. Levdistance is the levenshtein distance between the output of the
lossified grammar and the original input string, it is a kind of measure of the
fidelity of our algorithms but doesn't take into account how well patterns were
preserved. For the rep-n tests, proportion is the proportion of the final
string produced that was `*'. The closer to 1/30~0.033, the more accurate the
lossifier was in recreating the original texture of the input.

The columns of the data are prefixed by S or T depending on whether they are
measurements for \Similarity or \Cluster respectively.

@@

The actual compression ratio was respectable, with both \Similarity and \Cluster
coming in at nearly half the size of the \Sequitur CFG.  Setting higher values
for $\epsilon$ had diminishing returns for compression for larger and larger
file sizes, at the cost of drastically lowering the fidelity of the output.

@@

The differences between the two algorithms lie almost entirely in their
runtime. Both algorithms perform much worse on the less structued gen-n data,
we believe because simplifications are made more incrementally and the grammar
does not get dramatically simpler after a few iterations of our algorithm like
it does for the rep-n data.  Unfortunately the \Similarity algorithm tends to
consistently incur an overhead of about $O(\sqrt{n})$ on top of the runtime for
the \Cluster algorithm. Because of this we dismiss the \Similarity algorithm as
unscalable and do not test it further.

@@

Though we were ultimately not able to come up with a good way to quantify how
much of the structure of a text was preserved, we have included the final
output strings we obtain after running the \Cluster algorithm with $\epsilon
= .4$ in the Appendices. We believe these tests serve as a good (subjective)
summary of the strengths and weaknesses of our algorithm in compressing data
while preserving its character.

The first piece of text is from the gen-n dataset again, the first 256000
characters of the book of Genesis. As listed in table \ref{simclus_g_t}, the
compressed grammar was size 5540 compared with \Sequitur's size of 8060.
Without knowledge of English spelling, the \Cluster algorithm produces output
that is hard to read. However, key words are clearly recognizable, especially
important words such as `God', and the overall structure in the book of Genesis
is retained.
